Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2012-03-01 pkg/paper/Paper/EmpEc.Rnw
renamed JAE.Rnw to EmpEc.Rnw for new version
employ JSS style (instead of mimicking JAE style)

add diagnostics table (HAC, GOF, autocorrelation)



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@125 b679b2f7-6539-45d0-a605-5ddcb976cde9

\documentclass[10pt,a4paper,twoside]{article}

%% additional packages
\usepackage{amsfonts,amstext,amsmath,a4wide,graphicx,color,thumbpdf}
\usepackage{hyperref}
\newcommand{\argmax}{\operatorname{argmax}\displaylimits}

@@

%% BibTeX settings
\usepackage[authoryear,round]{natbib}
\bibliographystyle{jae}
\bibpunct{(}{)}{,}{a}{,}{,}
\newcommand{\doi}[1]{\href{http://dx.doi.org/#1}{\normalfont\texttt{doi:#1}}}

%% markup commands for code/software
\let\code=\texttt
\let\pkg=\textbf
\let\proglang=\textsf
\newcommand{\file}[1]{`\code{#1}'}
\newcommand{\email}[1]{\href{mailto:#1}{\normalfont\texttt{#1}}}

%% paragraph formatting
\renewcommand{\baselinestretch}{1.52}

%% mimic JAE style
\renewcommand{\section}{\secdef \mysec \mysecnn}
\newcommand{\mysec}[2][default]{\vspace{1.7\baselineskip}%
  \pdfbookmark[1]{#1}{Section.#1}%
  \refstepcounter{section}%
  \centerline{\large \thesection. \uppercase{#1}} \vspace{.5\baselineskip}}
\newcommand{\mysecnn}[1]{\vspace{1.7\baselineskip}%
  \centerline{\large #1} \vspace{.5\baselineskip}}
\renewcommand{\abstractname}{\normalfont SUMMARY}
\renewcommand{\refname}{REFERENCES}
\renewcommand{\thetable}{\Roman{table}}

%% page header (and currently no footer)
\usepackage{fancyhdr}
\setlength{\headheight}{15pt}
\renewcommand{\headrulewidth}{0pt}
\pagestyle{fancy}
\thispagestyle{plain}
\fancyhf{}
\fancyhead[LE,RO]{\thepage}
\fancyhead[CE]{{\normalsize \uppercase{T.~Windberger and A.~Zeileis}}}
\fancyhead[CO]{{\normalsize \uppercase{Structural Breaks in Inflation Dynamics within the EMU}}} 
% \fancyfoot[LO,LE]{\small Copyright {\copyright} 20XX John Wiley \& Sons, Ltd.}
\fancypagestyle{plain}{
\fancyhf{}
% \fancyfoot[LO,LE]{\small This is a preprint of an article published in %
%   \textit{Journal of Applied Econometrics}, \textbf{XX}(X), XXX--XXX. \\%
%   Copyright {\copyright} 20XX John Wiley \& Sons, Ltd. \doi{10.1002/jae.XXXX} }
}

%% title information
\title{\bf Structural Breaks in Inflation Dynamics within the European Monetary Union}
\author{\hfill Thomas Windberger \hfill Achim Zeileis\thanks{Correspondence
to: Achim Zeileis, Department of Statistics,  Universit\"at Innsbruck,
Universit\"atsstr.~15, A-6020 Innsbruck, Austria. Tel: +43/512/507-7103. Fax: +43/512/507-2851.
E-mail: \email{Achim.Zeileis@R-project.org}} \hfill \hfill \\
{\small \it Department of Statistics, Universit\"at Innsbruck, Austria}}
\date{}

% hyperref setup
\definecolor{Red}{rgb}{0.5,0,0}
\definecolor{Blue}{rgb}{0,0,0.5}
\hypersetup{%
  pdftitle = {Structural Breaks in Inflation Dynamics within the European Monetary Union},
  pdfsubject = {},
  pdfkeywords = {inflation rate, structural break, EMU, generalized logistic distribution},
  pdfauthor = {Thomas Windberger, Achim Zeileis},
  %% change colorlinks to false for pretty printing
  colorlinks = {true},
  linkcolor = {Blue},
  citecolor = {Blue},
  urlcolor = {Red},
  hyperindex = {true},
  linktocpage = {true},
}

@@

\maketitle

\begin{abstract}
  To assess the effects of the EMU on inflation rate dynamics of
  its member states, the inflation rate series for 21~European countries
  are investigated for structural changes. To capture changes in mean,
  variance, and skewness of inflation rates, a generalized logistic
  model is adopted and complemented with structural break tests and
  breakpoint estimation techniques. These reveal considerable differences
  in the patterns of inflation dynamics and the structural changes
  therein. Overall, there is a convergence towards a lower mean inflation rate
  with reduced skewness, but it is accompanied by an increase in variance.
\end{abstract}

\noindent {\bf Keywords:} inflation rate, structural break, EMU, generalized logistic distribution.

@@

for(i in seq_along(nam)) {
  load(file.path(root_dir, "Analysis", paste(nam[i], "rda", sep = ".")))
  results[[i]] <- list(
    data = x,
    model0 = x_gf,
    efp = x_efp,
    nbreaks = x_nbreaks,
    breakpoints = x_bp,
    model = x_rf,
    erm = erm[i],
    erm2 = erm2[i],
    eur = eur[i])
}
@

@@

\renewcommand{\baselinestretch}{1.15}
\begin{table}[p!]
\begin{center}
\footnotesize
\begin{tabular}{llrrrlll}
\hline
Country & Segment & Mean & Variance & Skewness & ERM & ERM~II & Euro \\
\hline

@@

\caption{\label{tab:results1} Estimated breakpoints, fitted moments of generalized logistic
distribution per segment, and ERM information for all countries under investigation.}
\end{center}
\end{table}
\renewcommand{\baselinestretch}{1.52}

@@

\renewcommand{\baselinestretch}{1.15}
\begin{table}[t!]
\begin{center}
\begin{tabular}{llrrrr}
\hline
Moment & Quartile & \Sexpr{yrs[1]} & \Sexpr{yrs[2]} & \Sexpr{yrs[3]} & \Sexpr{yrs[4]}\\
\hline

@@

\end{tabular}
\caption{\label{tab:results2} Aggregation of estimated moments from Table~\ref{tab:results1} across countries in
Jan~\Sexpr{yrs[1]}, \Sexpr{yrs[2]}, \Sexpr{yrs[3]}, and \Sexpr{yrs[4]}. Dots signal that the moments have not
changed from the previous time point, i.e., that no breakpoint was estimated in the corresponding time period.}
\end{center}
\end{table}
\renewcommand{\baselinestretch}{1.52}

@@

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[t!]
\begin{center}
<<glmean-plot, fig=TRUE, height=5, width=7.5>>=
  gray_red <- rgb(c(0.2, 0.8), c(0.2, 0), c(0.2, 0), alpha = 0.3)
  gray_red1 <- rgb(c(0.2, 0.8), c(0.2, 0), c(0.2, 0))
  plot(hicps, plot.type = "single", lwd = 1.5, col = gray_red[1],
    xlab = "Time", ylab = "Monthly inflation rates")
  for(i in 1:ncol(hicps)) lines(hicp_glmean[,i], lwd = 1.5, col = gray_red[2])
@
\caption{\label{fig:glmean} Observed monthly inflation rates (gray) and estimated means from
12-month rolling GL fits (red) over time for all countries.}
\end{center}
\end{figure}
\setkeys{Gin}{width=\textwidth}

@@

\section*{COMPUTATIONAL DETAILS}

>>>>

\documentclass[nojss]{jss}
\usepackage{amsfonts,amstext,amsmath,thumbpdf}
\newcommand{\argmax}{\operatorname{argmax}\displaylimits}

@@

\title{Structural Breaks in Inflation Dynamics within the European Monetary Union}

\author{Thomas Windberger\\Universit\"at Innsbruck \And
        Achim Zeileis\\Universit\"at Innsbruck}
\Plainauthor{Thomas Windberger, Achim Zeileis}

\Abstract{
  To assess the effects of the EMU on inflation rate dynamics of
  its member states, the inflation rate series for 21~European countries
  are investigated for structural changes. To capture changes in mean,
  variance, and skewness of inflation rates, a generalized logistic
  model is adopted and complemented with structural break tests and
  breakpoint estimation techniques. These reveal considerable differences
  in the patterns of inflation dynamics and the structural changes
  therein. Overall, there is a convergence towards a lower mean inflation rate
  with reduced skewness, but it is accompanied by an increase in variance.
}

\Keywords{inflation rate, structural break, EMU, generalized logistic distribution}

\Address{
  Thomas Windberger, Achim Zeileis\\
  Department of Statistics\\
  Faculty of Economics and Statistics\\
  Universit\"at Innsbruck\\
  Universit\"atsstra{\ss}e 15\\
  6020 Innsbruck, Austria\\
  E-mail: \email{Thomas.Windberger@student.uibk.ac.at}, \email{Achim.Zeileis@R-project.org} \\
  Telephone: +43/12/507-7103 \\
  Fax: +43/512/507-2851 \\
}

@@

@@

for(i in seq_along(nam)) {
  load(file.path(root_dir, "Analysis", paste(nam[i], "rda", sep = ".")))
  results[[i]] <- list(
    data = x,
    model0 = x_gf,
    efp = x_efp,
    efp_hac = gefp(x_gf, fit = NULL, vcov = vcovHAC), #FIXME# added in revision for EmpEc, should be moved to 'Analysis'
    nbreaks = x_nbreaks,
    breakpoints = x_bp,
    model = x_rf,
    erm = erm[i],
    erm2 = erm2[i],
    eur = eur[i])
}
@

@@

\begin{table}[p!]
\centering
\footnotesize
\begin{tabular}{llrrrlll}
\hline
Country & Segment & Mean & Variance & Skewness & ERM & ERM~II & Euro \\
\hline

@@

\caption{\label{tab:results1} Estimated breakpoints, fitted moments of generalized logistic
distribution per segment, and ERM information for all countries under investigation.}
\end{table}

<<results-diagnostics>>=
names(results)[names(results) == "CzechRepublic"] <- "Czech Rep."
format_diag <- function(i, digits = 3) {
  pround <- function(x) ifelse(abs(x) < 0.001, "$< 0.001$",
    paste("$", format(round(x, digits = digits), nsmall = digits), "$", sep = ""))

  rval <- cbind("", "", names(results[[i]]$model), "", "", "")
  n <- NROW(rval)
  if(n <= 1) rval <- rbind(rval, "")

  rval[1,   1] <- if(is.character(i)) i else names(results)[i]
  rval[1,   2] <- pround(sctest(results[[i]]$efp,     functional = supLM(0.1))$p.value)
  rval[2,   2] <- pround(sctest(results[[i]]$efp_hac, functional = supLM(0.1))$p.value)
  rval[1:n, 4] <- pround(sapply(results[[i]]$model, function(obj) summary(obj)$chisq.test$p.value))
  rval[1:n, 5] <- pround(sapply(results[[i]]$model, function(obj) acf(obj$x, plot = FALSE)$acf[2]))
  rval[1:n, 6] <- pround(sapply(results[[i]]$model, function(obj) Box.test(obj$x, type = "Ljung-Box")$p.value))

  rval <- apply(rval, 1, paste, collapse = " & ")
  rval <- paste(rval, "\\\\")
  rval[length(rval)] <- paste(rval[length(rval)], "\\hline")
  return(rval)
}
@

\begin{table}[p!]
\centering
\footnotesize
\begin{tabular}{lrlrrr}
\hline
Country & sup$\mathit{LM}$ (standard/HAC) & Segment & $\chi^2$ GOF & AR(1) & Ljung-Box \\
\hline

\noalign{\medskip}
\multicolumn{6}{l}{\emph{No change}} \\
\hline
<<diagnostics-table1, results=tex>>=
writeLines(unlist(lapply(c(2, 4, 8, 14), format_diag)))
@

\noalign{\medskip}
\multicolumn{6}{l}{\emph{Phase~1 convergence}} \\
\hline
<<diagnostics-table2, results=tex>>=
writeLines(unlist(lapply(c(1, 6, 7), format_diag)))
@

\noalign{\medskip}
\multicolumn{6}{l}{\emph{Phase~2 convergence}} \\
\hline
<<diagnostics-table3, results=tex>>=
writeLines(unlist(lapply(c(9, 12, 13, 16), format_diag)))
@

\noalign{\medskip}
\multicolumn{6}{l}{\emph{Financial crisis}} \\
\hline
<<diagnostics-table4, results=tex>>=
writeLines(unlist(lapply(c(11, 19), format_diag)))
@

\noalign{\medskip}
\multicolumn{6}{l}{\emph{Non-euro}} \\
\hline
<<diagnostics-table5, results=tex>>=
writeLines(unlist(lapply(c(20, 21), format_diag)))
@

\noalign{\medskip}
\multicolumn{6}{l}{\emph{Eastern countries}} \\
\hline
<<diagnostics-table6, results=tex>>=
writeLines(unlist(lapply(c(3, 5, 10, 15, 17, 18), format_diag)))
@

\noalign{\medskip}
\multicolumn{6}{l}{\emph{Euro area}} \\
\hline
<<diagnostics-table6, results=tex>>=
writeLines(unlist(lapply(22, format_diag)))
@

\caption{\label{tab:diag} Diagnostics for model fits in each country. Two sup$\mathit{LM}$ test $p$~values
  of the full-sample model with and without HAC correction and segment-specific assessments of the model
  fit: $\chi^2$ goodness-of-fit (GOF) $p$~value; autocorrelation at lag~1; Ljung-Box $p$~value.}
\end{table}

@@

\begin{table}[t!]
\centering
\begin{tabular}{llrrrr}
\hline
Moment & Quartile & \Sexpr{yrs[1]} & \Sexpr{yrs[2]} & \Sexpr{yrs[3]} & \Sexpr{yrs[4]}\\
\hline

@@

\end{tabular}
\caption{\label{tab:results2} Aggregation of estimated moments from Table~\ref{tab:results1} across countries in
Jan~\Sexpr{yrs[1]}, \Sexpr{yrs[2]}, \Sexpr{yrs[3]}, and \Sexpr{yrs[4]}. Dots signal that the moments have not
changed from the previous time point, i.e., that no breakpoint was estimated in the corresponding time period.}
\end{table}

@@

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[t!]
\centering
<<glmean-plot, fig=TRUE, height=5, width=7.5>>=
  gray_red <- rgb(c(0.2, 0.8), c(0.2, 0), c(0.2, 0), alpha = 0.3)
  gray_red1 <- rgb(c(0.2, 0.8), c(0.2, 0), c(0.2, 0))
  plot(hicps, plot.type = "single", lwd = 1.5, col = gray_red[1],
    xlab = "Time", ylab = "Monthly inflation rates")
  for(i in 1:ncol(hicps)) lines(hicp_glmean[,i], lwd = 1.5, col = gray_red[2])
@
\caption{\label{fig:glmean} Observed monthly inflation rates (gray) and estimated means from
12-month rolling GL fits (red) over time for all countries.}
\end{figure}
\setkeys{Gin}{width=\textwidth}

@@

\section*{Computational details}
